Parsing Postfix log files

### 2007-11-25 doc/logparser.tex
Finished the explanation of the logs2regexs algorithm.

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, continuing with detecting and dealing
with overlapping rules, and the problems they can cause.  An example rule
and a line it would match are provided, plus a description of how the
fields in the rule are used when matching and performing the action (the
details of the table containing the rules have already been described in
section~\refwithpage{rule attributes},).  This section continues with a
discussion of rule efficiency concerns, referring to the graphs in
section~\refwithpage{graphs}, and finishes with an explanation of how to
add or edit rules, including the algorithm used to generate a new \regex{}
from unparsed lines.

@@

Perl compiles the original \regex{} into an internal representation,
optimising the \regex{} to improve the speed of matching, but this
compilation and optimisation takes CPU time.  Perl automatically caches
static \regexs{}, but dynamic \regexs{} need to be explicitly compiled and
cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
times with and without caching the \regex{}.  Caching the compiled
\regexs{} is obviously far more efficient; graph~\refwithpage{normal regex
vs discarded regex factor} shows the percentage execution time increase
when not caching each \regex{}.

@@

The logs produced by Postfix differ from installation to installation, as
administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \RBL{}
services, policy servers, or custom rejection messages.  To facilitate
this, the parser's design separates parsing rules from parsing actions:
adding new actions is difficult, but adding new rules to parse new
rejection messages is trivial (and occurs much more frequently).  The
implementation provides a program to ease the process of creating new rules
from unparsed lines, based on the algorithm developed by Risto
Vaarandi~\cite{risto-vaarandi} for \SLCT{}~\cite{slct-paper}.  The
differences between the two algorithms will be outlines as part of the
general explanation below.

The core of the algorithm was developed by Risto Vaarandi, and is quite
simple: log lines are generally created by substituting variable words into
a fixed pattern, and analysis of the frequency with which each word occurs
determines whether the word is fixed or variable.  This classification can
be used to group similar log lines and generate a \regex{} to match each
group of lines.

@@

    \item [Pre-process the file]

        In the original algorithm the preprocessing stage was

    \item [Calculate the word frequency]

    \item [XXX]

    \item [Build regexs]

@@

@@

    \item This section finished with an explanation of how to add or edit
        rules, including the algorithm used to generate a new \regex{} from
        unparsed lines.

@@

% This is very likely to move around as the flow chart is as big as
% possible, so if any preceding text changes the flow chart may end up on
% the next page, which may not be a bad thing, maybe.  What woould be a bad
% thing is the \clearpage seems to cause a blank page sometimes, if the
% image is just slightly too big.
\showgraph{build/logparser-flow-chart.ps}{Parser flow chart}{flow chart
image}
\clearpage

@@

    \item [matches] The fields in the line captured by the rule's regex.

@@

>>>>

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, continuing with detecting and dealing
with overlapping rules, and the problems they can cause.  An example rule
and a line it would match are provided, plus a description of how the
fields in the rule are used when matching and performing the action (the
details of the table containing the rules have already been described in
section~\refwithpage{rule attributes},).  This section continues with a
discussion of rule efficiency concerns, referring to the graphs in
section~\refwithpage{graphs}, and finishes with an explanation of the
algorithm used to generate a new \regex{} from unparsed lines.

@@

Perl compiles the original \regex{} into an internal representation,
optimising the \regex{} to improve the speed of matching, but this
compilation and optimisation takes CPU time; far more CPU time, in fact,
than the actual matching takes.  Perl automatically caches static
\regexs{}, but dynamic \regexs{} need to be explicitly compiled and cached.
Graph~\refwithpage{normal regex vs discard regex} shows execution times
with and without caching the \regex{}.  Caching the compiled \regexs{} is
obviously far more efficient; graph~\refwithpage{normal regex vs discarded
regex factor} shows the percentage execution time increase when not caching
each \regex{}.

@@

The logs produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \RBL{}
services, policy servers, or custom rejection messages.  To facilitate easy
parsing of new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions is difficult, but adding new rules to
parse new rejection messages is trivial (and also occurs much more
frequently).  The implementation provides a program to ease the process of
creating new rules from unparsed lines, based on the algorithm developed by
Risto Vaarandi~\cite{risto-vaarandi} for his \SLCT{}~\cite{slct-paper}.
The differences between the two algorithms will be outlined as part of the
general explanation below.

The core of the algorithm is quite simple: log lines are generally created
by substituting variable words into a fixed pattern, and analysis of the
frequency with which each word occurs can be used to determine whether the
word is fixed or variable.  This classification can be used to group
similar log lines and generate a \regex{} to match each group of lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained while writing rules and performs a large number of
        substitutions on the input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        which the parser will expand when it loads the rule (see the
        \regex{} entry in section~\refwithpage{rule attributes}).  The
        purpose of this phase is to utilise the existing knowledge to
        create more accurate \regexs{}.  The new log lines are written to a
        temporary file, which all subsequent stages use instead of the
        original input file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash from a small range of values for each word
        of each line, incrementing a counter for each hash.  The counters
        will later be used to filter out words: if the word's hash does not
        have a high frequency, the word itself cannot have a high
        frequency, and there is no need to maintain a counter for it,
        reducing the number of counters and thus the programme's memory
        consumption.

    \item [Calculate word frequencies]  The position of words within a line
        is important: a common word does not indicate similarity between
        lines unless it occupies the same position within both
        lines.\footnote{If a variable term within a line contains spaces,
        it will appear to the algorithm as two words rather than one.  This
        will alter the position of subsequent words, so a word occurring in
        different positions in two log lines may indicate similarity, but
        the algorithm does not attempt to deal with this possibility.}  The
        algorithm maintains a counter for each \textit{(word, word's
        position within the line)\/} tuple, incrementing it each time that
        word is occurs in that position.

        The original algorithm only maintains counters for words whose hash
        result has a high frequency from the previous phase; this reduces
        the number of counters maintained by the algorithm, reducing the
        memory requirements of the algorithm.  The modified algorithm omits
        this check because the majority of unique or infrequently occurring
        words will have been substituted with keywords during the first
        phase, vastly reducing the number of tuples to maintain counters
        for.

    \item [Classify words based on their frequency]  The frequency of each
        \textit{(word, word's position within the line)\/} tuple is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all lines is generally a good starting point) it
        is classified as a fixed word, otherwise it is classified as a
        variable term.  Variable terms are replaced by \texttt{.*}, which
        means to match zero or more of any character.  The words are
        reassembled to produce a \regex{} matching the line, and a counter
        is maintained for each \regex{}.

    \item [Build regexs]  Contiguous sequences of \texttt{.*} in the newly
        reassembled \regexs{} are collapsed to a single \texttt{.*}; any
        resulting duplicate \regexs{} are combined, and their counters
        summed.  If the frequency of a \regex{} is lower than the threshold
        supplied by the user the \regex{} is discarded.\footnote{This is a
        second, independent threshold, but once again 1\% of input lines is
        a good starting point; obviously the threshold depends on the
        number and type of input lines.}  The new \regexs{} are printed for
        the user to add to the database, either as new rules or merged into
        the \regexs{} of existing rules; the counter for each \regex{} is
        also printed, giving the user an indication of how many of the
        input lines that \regex{} should match.  Discarding \regexs{} will
        result in some of the input lines not being matched; this utility
        should be run again once the unmatched lines have been isolated by
        running the parser, including the new \regexs{}, with the same
        input which produced the original set of unparsed lines.

@@

A second utility is also provided which reads a list of new \regexs{} and
the input given to the first utility.  It tries to match each input line
against each \regex{}, counting the number of lines which match each
\regex{}, warning the user if an input line is matched by more than one
\regex{}, and additionally warning if an input line is not matched by any
\regex{}.  It displays a summary of how many input lines each \regex{}
matched, comparing it to the expected number of matches; this provides the
user with an easy method of checking if the \regexs{} produced by the first
utility are correctly matching the input lines they are based upon.  A
future version of this utility will also group input lines by \regex{}, so
the user can tweak the \regexs{} if required.

@@

    \item This section finished with an explanation of the algorithm used
        to generate a new \regex{} from unparsed lines.

@@

XXX SPLIT THIS IMAGE INTO TWO\@; IT'S UNREADABLE NOW, IT'S SO BIG\@.

% This is very likely to move around as the flow chart is as big as
% possible, so if any preceding text changes the flow chart may end up on
% the next page, which may not be a bad thing, maybe.  What would be a bad
% thing is the \clearpage seems to cause a blank page sometimes, if the
% image is just slightly too big.
\showgraph{build/logparser-flow-chart.ps}{Parser flow chart}{flow chart
image}
\clearpage

@@

    \item [matches] The fields in the line captured by the rule's \regex{}.

@@

\glossary{name={hash},description={
    A hashing function transforms a string of characters to a number.
    There are many possible uses for the resulting number: a common usage
    is to maintain a data structure indexed by strings in an efficient
    manner.  A full description is beyond the scope of this paper, further
    information can be found at~\cite{hash-functions}.
}}
